---
title: "Sampling distributions for reaction time data and a warning about data-driven simulations"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  # html_document:
  github_document:
    # dev: CairoPDF
    # fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

# Dependencies
```{r message=FALSE, warning=FALSE}
library(ggplot2)
source("./functions/theme_gar.txt") # define ggplot theme
source("./functions/trimpval.txt") # get one-sample t-test on trimmed means
source("./functions/functions.txt")
library(tibble)
library(cowplot)
library(rogme) # this package is not on CRAN, follow these steps to install it:
# install.packages("devtools")
# devtools::install_github("GRousselet/rogme")
library(beepr)
```

```{r}
sessionInfo()
```

Before we conduct an experiment, we decide on a number of trials per condition, and a number of participants, and then we hope that whatever we measure comes close to the population values we want to estimate. Using a large dataset, we can perform data-driven simulations to illustrate how close we can get to the truth.

We use data from the [French Lexicon Project](https://sites.google.com/site/frenchlexicon/results).
Click on "French Lexicon Project trial-level results with R scripts.zip".
The `.RData` dataset was created by applying the script `getflprtdata.Rmd` available on [GitHub](https://github.com/GRousselet/rogme/tree/master/data-raw).

```{r}
# get data - tibble = `flp`
flp <- rogme::flp
# columns =
#1 = participant
#2 = rt
#3 = acc = accuracy 0/1
#4 = condition = word/non-word

# Determine population trimmed mean of trimmed mean differences
tmp <- tapply(flp$rt, list(flp$participant, flp$condition), mean, trim = 0.2)
pop.tm.diff <- mean(tmp[,2]-tmp[,1], trim = 0.2)
pop.np <- length(unique(flp$participant))
```

N = `r length(unique(flp$participant))` participants.

# Illustrate random sample of participants

## Create data frame
```{r}
set.seed(3)
# participants' indices
p.list <- as.numeric(as.character(unique(flp$participant)))
np <- 20
rsamp <- sample(p.list, np, replace = FALSE)
df <- flp[flp$participant %in% rsamp, ]
df$participant <- as.character(df$participant)
df$participant <- factor(df$participant, levels=unique(df$participant))
#levels(df$condition)
levels(df$condition) <- c("Word", "Non-Word")
```

## Make figure
```{r, fig.height=4, fig.width=10}
p <- ggplot(df, aes(x = rt)) + theme_gar +
  geom_line(stat = "density", aes(colour=participant), linewidth = 0.5) + 
  scale_color_viridis_d() + 
  scale_x_continuous(breaks=seq(0,2000,500), minor_breaks = seq(0, 2000, 250)) +
  coord_cartesian(xlim = c(0, 2000)) +
  theme(legend.position = "none") + 
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(x = "Reaction times (ms)", y = "Density") +
  facet_grid(. ~ condition)
p
p.20 <- p
```

The distributions are positively skewed, as expected for RT data, and participants tend to be slower in the non-word condition compared to the word condition. Usually, a single number is used to summarise each individual RT distribution. From 1000 values to 1, that’s some serious data compression! In psychology, the mean is often used, but here we use the 20% trimmed mean, which gives a better indication of the location of the typical observation and protects against the influence of outliers. 

# Summary results for all participants

Here is the distribution across participants of trimmed mean RT for the word and non-word conditions and their differences.

## Compute trimmed means
```{r}
# get data: mean RT for every participant
tmres <- tapply(flp$rt, list(flp$participant, flp$condition), mean, trim = 0.2)
summary(tmres)
```

## Illustrate results

All distributions are skewed.

```{r, fig.height=4, fig.width=10}

# create data frame
df <- tibble(x = c(as.vector(tmres),tmres[,2]-tmres[,1]),
             Condition = rep(c("Word", "Non-Word", "Difference"), each = pop.np)
)
df$Condition <- as.character(df$Condition)
df$Condition <- factor(df$Condition, levels=unique(df$Condition))

facet_bounds <- read.table(header=TRUE,
text=                           
"Condition xmin xmax breaks
Word 0 1600 250
Non-Word 0 1600 250
Difference -100 500 100",
stringsAsFactors=FALSE)

ff <- with(facet_bounds,
           data.frame(x=c(xmin,xmax),
                      Condition=c(Condition,Condition)))

ff$Condition <- keeporder(ff$Condition)

# make plot
p <- ggplot(df, aes(x = x)) + theme_gar + 
  geom_line(stat = "density", linewidth = 1.5) + 
  labs(x = "20% trimmed mean reaction times (ms)", y = "Density") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_grid(. ~ Condition, scales = "free") + 
  geom_blank(data=ff)
p
p.alltm <- p
```

`r round(mean(tmres[,2]-tmres[,1]>0)*100, digits=1)`% of participants had a positive difference, meaning that they tended to be faster in the Word than the Non-Word condition.

# Hierarchical sampling: effect

We sample participants and trials with replacement. For each condition, participant and trial sample size, we compute the 20% trimmed mean. Then, for each condition and participant sample size we compute the group 20% trimmed mean and the *p* value based on a t-test for trimmed means.

```{r, eval = FALSE}
set.seed(22222)
p.list <- as.numeric(as.character(unique(flp$participant)))
nsim <- 5000
np.seq <- c(10, 20, 50) # number of participants
maxNP <- max(np.seq)
nt.seq <- seq(50, 250, 50) # number of trials
maxNT <- max(nt.seq)

# intermediate results with single trials
diff.all <- matrix(NA, nrow = maxNP, ncol = length(nt.seq))

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(nt.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  if(iter %% 100 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  # Sample participants with replacement
  bootid <- sample(p.list, maxNP, replace = TRUE)
  # Sample trials with replacement 
  for(P in 1:maxNP){ # for each bootstrap participant
    boot.w <- sample(flp$rt[flp$participant %in% bootid[P] & flp$condition == "word"],
                     maxNT, replace = TRUE)
    boot.nw <- sample(flp$rt[flp$participant %in% bootid[P] & flp$condition == "non-word"],
                      maxNT, replace = TRUE)
    
    # compute individual differences between trimmed means
    for(T in 1:length(nt.seq)){ # for each sample size
      # compute trimmed means
      diff.all[P,T] <- mean(boot.nw[1:nt.seq[T]], trim = 0.2) - mean(boot.w[1:nt.seq[T]], trim = 0.2)
    }
  }
  
  # compute group trimmed means and p values 
  for(NP in 1:length(np.seq)){
    for(NT in 1:length(nt.seq)){
      sim.res[iter, NP, NT] <- mean(diff.all[1:np.seq[NP],NT], trim = 0.2)
      sim.res.pval[iter, NP, NT] <- trimpval(diff.all[1:np.seq[NP],NT], tr = 0.2)
    }
  }
}

save(
  nsim,
  np.seq,
  nt.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res.RData"  
)

beep(8)
```

## Illustrate results

### Panel 1
Original sampling distributions.
```{r, fig.height=4, fig.width=10}
load(file = "./data/flp_sim_res.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
)

levels(df$np) <- c("10 participants", "20 participants", "50 participants")

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = pop.tm.diff) +
  geom_line(stat = "density", linewidth = 0.75) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p.diff1 <- p
p
```

### Panel 2
Plot sampling distributions conditional on statistical significance.  
```{r, fig.height=4, fig.width=10, warning=FALSE}
alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res[v.sim.res.pval>alpha] <- NA

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
)

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = pop.tm.diff) +
  geom_line(stat = "density", linewidth = 0.75) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

The vertical line in each panel marks the population 20% trimmed mean.
Here, conditioning on $p\leq0.05$ has very little effect on the sampling distributions because the population effect size is so large.

# Hierarchical sampling: no effect

In this situation, trials are sampled with replacement from the two conditions merged together, so that on average, there is no effect.

```{r, eval = FALSE}
set.seed(22222)
p.list <- as.numeric(as.character(unique(flp$participant)))
nsim <- 5000
np.seq <- c(10, 20, 50) # number of participants
maxNP <- max(np.seq)
nt.seq <- seq(50, 250, 50) # number of trials
maxNT <- max(nt.seq)

# intermediate results with single trials
diff.all <- matrix(NA, nrow = maxNP, ncol = length(nt.seq))

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(nt.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  if(iter %% 100 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  # Sample participants with replacement
  bootid <- sample(p.list, maxNP, replace = TRUE)
  # Sample trials with replacement 
  for(P in 1:maxNP){ # for each bootstrap participant
    all.trials <- flp$rt[flp$participant %in% bootid[P] & 
                           (flp$condition == "word" | flp$condition == "non-word")]
    boot.w <- sample(all.trials, maxNT, replace = TRUE)
    boot.nw <- sample(all.trials, maxNT, replace = TRUE)
    
    # compute individual differences between trimmed means
    for(T in 1:length(nt.seq)){ # for each sample size
      # compute trimmed means
      diff.all[P,T] <- mean(boot.nw[1:nt.seq[T]], trim = 0.2) - mean(boot.w[1:nt.seq[T]], trim = 0.2)
    }
  }
  
  # compute group trimmed means and p values 
  for(NP in 1:length(np.seq)){
    for(NT in 1:length(nt.seq)){
      sim.res[iter, NP, NT] <- mean(diff.all[1:np.seq[NP],NT], trim = 0.2)
      sim.res.pval[iter, NP, NT] <- trimpval(diff.all[1:np.seq[NP],NT], tr = 0.2)
    }
  }
}

save(
  nsim,
  np.seq,
  nt.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res_noeffect.RData"  
)

beep(8)
```

## Illustrate results

### Panel 1
```{r, fig.height=4, fig.width=10}
load(file = "./data/flp_sim_res_noeffect.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
)

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = 0) + # long-run effect size
  geom_line(stat = "density", linewidth = 0.75) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

### Panel 2
Same as panel 1, now only plotting differences if test was significant.
```{r, fig.height=4, fig.width=10, warning=FALSE}
alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res[v.sim.res.pval>alpha] <- NA

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
)

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = 0) +
  geom_line(stat = "density", linewidth = 0.75) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

Results conditioned on $p\leq0.05$ are symmetric around zero. With 10 participants and 50 trials per condition, the median of the absolute group differences is `r round(median(abs(df$x[df$np=="10" & df$nt=="50"]), na.rm = TRUE), digits=1)` ms, and among the experiments reporting significant results, 10% report differences at least as large as `r round(quantile(abs(df$x[df$np=="10" & df$nt=="50"]), na.rm = TRUE, probs=0.9, type=8), digits=1)` ms. With 50 participants and 200 trials per condition, the median of the absolute group differences is `r round(median(abs(df$x[df$np=="50" & df$nt=="200"]), na.rm = TRUE), digits=1)` ms, and among the experiments reporting significant results, 10%  report differences at least as large as `r round(quantile(abs(df$x[df$np=="50" & df$nt=="200"]), na.rm = TRUE, probs=0.9, type=8), digits=1)` ms.

### Merged panels using facetting
```{r, eval = FALSE, warning=FALSE}
load(file = "./data/flp_sim_res_noeffect.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res.raw <- as.vector(sim.res)
v.sim.res.cond <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res.cond[v.sim.res.pval>alpha] <- NA

df <- tibble(x = c(v.sim.res.raw, v.sim.res.cond),
             np = factor(rep(rep(rep(np.seq, each = nsim), length(nt.seq)),2)),
             nt = factor(rep(rep(nt.seq, each = nsim * length(np.seq)),2)),
             cond = factor(rep(c("Raw data","Cond. data"), 
                               each = nsim * length(np.seq) * length(nt.seq)))
)

levels(df$np) <- c("10 participants", "20 participants", "50 participants")
df$cond <- keeporder(df$cond)

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = 0) + # long-run effect size
  geom_line(stat = "density", linewidth = 0.75) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(rows = vars(cond), cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p.diff.noeffect <- p
p
```

## Summary figure
```{r, eval=FALSE, warning=FALSE}

cowplot::plot_grid(p.20, 
                   p.alltm, 
                   p.diff1, 
                   p.diff.noeffect + theme(legend.position = "none"),
                   labels = c("A", "B", "C", "D"),
                   ncol = 1,
                   nrow = 4,
                   rel_heights = c(1, 0.75, 0.75, 1), 
                   label_size = 20,
                   align = 'v',
                   axis = 'l',
                   hjust = -0.5,
                   scale = 1)

# save figure
ggsave(filename=('./figures/fig_flp.tiff'),width=13,height=20,dpi=300)
```

# Effect of population size on sampling distributions

As described in [Burns, Fracasso & Rousselet (2025)](https://www.nature.com/articles/s41598-025-89257-w), data-driven simulations are affected by the relative difference between the population size and the sample size. As sample sizes get closer to the population size, estimation bias increases, and the sign of the bias depends on the effect size in the population. Let's illustrate the sampling distributions.

## Sampling distributions

For illustration, we consider sampling distributions for different combinations of participant population sizes and sample sizes, using all the trials available for each participant. We also calculate one-sample t-test for 20% trimmed means, with the null value set to 60 ms.

Concretely, for each participant, we calculate the 20% trimmed mean across all trials for each condition, and save the difference between the two conditions. The full size population is then defined as the one-sample distribution of trimmed mean differences for all participants. 

In each simulation iteration, populations of sizes 50, 100, ..., 250 are created by sampling without replacement from the full sample size (959 participants). Then, for each population size, experiments are simulated by sampling 20, 50 or 100 participants with replacement. It might seem strange to sample with replacement 100 participants from a population of 50, but I've seen that type of over-sampling a few times. As we will see shortly, it is a bad idea. For each, we calculate the group 20% trimmed mean.

```{r, eval=FALSE}
diff <- tmres[,2]-tmres[,1]
set.seed(22222)
nsim <- 20000
np.seq <- c(20, 50, 100) # number of participants
maxNP <- max(np.seq)
npop.seq <- seq(50, 250, 50) # population size
maxNPOP <- max(npop.seq)

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(npop.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  
  if(iter %% 1000 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  
  # Sample participants without replacement to create population
  pop.sub <- sample(diff, maxNPOP, replace=FALSE)
  
  for(POP in 1:length(npop.seq)){
    
    # downsample to current population size
    # sample participants with replacement from population
    samp.max <- sample(pop.sub[1:npop.seq[POP]], maxNP, replace=TRUE) 
    
    for(P in 1:length(np.seq)){ # for each sample size
      sim.res[iter, P, POP] <- mean(samp.max[1:np.seq[P]], trim=0.2)
      sim.res.pval[iter, P, POP] <- trimpval(samp.max[1:np.seq[P]], tr=0.2, null.value=60)
    } # participants
  } # pop size
} # simulation iterations

save(
  nsim,
  np.seq,
  npop.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res_bias.RData"
)

beep(8)
```

### Results

```{r, fig.height=4, fig.width=10}
load(file = "./data/flp_sim_res_bias.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(npop.seq))),
             npop = factor(rep(npop.seq, each = nsim * length(np.seq)))
)

levels(df$np) <- c("20 participants", "50 participants", "100 participants")

p <- ggplot(df, aes(x = x, colour = npop)) + theme_gar +
      geom_vline(xintercept = pop.tm.diff) +
      geom_line(stat = "density", linewidth = 0.75) +
      theme(legend.position = "bottom",
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank(),
            panel.grid.minor.y = element_blank()) +
      guides(colour = guide_legend(override.aes = list(linewidth = 3),
                                   title = "Population Size")) +
      scale_colour_viridis_d(option = "plasma", 
                             begin = 0, end = 0.7) + 
      facet_grid(cols = vars(np), scales = "fixed") +
      labs(x = "Group 20% trimmed mean difference", y = "Density") +
      coord_cartesian(xlim = c(35, 150)) + 
      scale_x_continuous(breaks = c(50, 100, 150))
p.biasA <- p
p
```

As it should be, the spread of the sampling distributions varies inversely with the sample size, here the number of participants. Now, what is interesting, is that for a fixed sample size, we get broader sampling distributions for smaller populations, and the problem is worse if our samples are large relative to the population size -- here we see a larger difference between population sizes 50 and 250 when taking samples of 100 participants rather than 20 participants. This phenomenon is due to the presence, in some populations, of an over-representation of extreme values, which are themselves more likely to be picked up when sampling with replacement in a simulation. As a result, we get exaggerated tails, with important consequences for power analyses for instance. 

## Power

New simulation in which we explore more sample and population sizes.
Inference is on the population 20% trimmed mean using a modified one-sample t-test.

```{r, eval=FALSE}
diff <- tmres[,2]-tmres[,1]
set.seed(22222)
nsim <- 20000
np.seq <- seq(10, 120, 10) # number of participants
maxNP <- max(np.seq)
npop.seq <- seq(50, 300, 50) # population size
maxNPOP <- max(npop.seq)

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(npop.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  
  if(iter %% 1000 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  
  # Sample participants without replacement to create population
  pop.sub <- sample(diff, maxNPOP, replace=FALSE)
  
  for(POP in 1:length(npop.seq)){
    
    # downsample to current population size
    # sample participants with replacement from population
    samp.max <- sample(pop.sub[1:npop.seq[POP]], maxNP, replace=TRUE) 
    
    for(P in 1:length(np.seq)){ # for each sample size
      sim.res[iter, P, POP] <- mean(samp.max[1:np.seq[P]], trim=0.2)
      sim.res.pval[iter, P, POP] <- trimpval(samp.max[1:np.seq[P]], tr=0.2, null.value=60)
    } # participants
  } # pop size
} # simulation iterations

save(
  nsim,
  np.seq,
  npop.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res_bias2.RData"
)

beep(8)
```

### Results
Here we plot power as a function of population size, separately for each sample size.
Alpha is set to the usual arbitrary value of 0.05.

```{r}
load("./data/flp_sim_res_bias2.RData")

df <- tibble(power = as.vector(apply(sim.res.pval <= 0.05, c(2,3), mean)),
             population = rep(npop.seq, each = length(np.seq)),
             Participant = factor(rep(np.seq, length(npop.seq)))
             )

p <- ggplot(df, aes(x=population, y=power, colour=Participant)) +
      theme_gar +
      geom_point() +
      geom_line() +
      scale_colour_viridis_d(option = "magma", 
                             begin = 0, end = 0.9) +
      scale_x_continuous(breaks = npop.seq) +
      scale_y_continuous(breaks = seq(0.1,0.9,0.1)) +
      labs(x = "Population Size", y = "Power")
p.biasB <- p
p
```

As expected, power depends on sample size, here the number of participants. But also notice how for each sample size, the population size has different effects. It is easier to see what is going on by focusing on the extremes: for the lowest sample size (n=10 participants), increasing the population size lowers power. In other words, for this large reaction time effect, doing a data-drive simulation using a small sample from a small population will tend to over-estimate statistical power. We get the opposite effect when we consider a larger sample size (n=120), as now a smaller population size leads to power under-estimation.

Alternatively, plot power as a function of sample size, separately for each population size.
Add reference line for 83% power.

```{r}
# load("./data/flp_sim_res_bias2.RData")

targpow <- .83 # target power -- 83 is a prime number

df <- tibble(power = as.vector(apply(sim.res.pval <= 0.05, c(2,3), mean)),
             population = factor(rep(npop.seq, each = length(np.seq))),
             participant = rep(np.seq, length(npop.seq))
             )

p <- ggplot(df, aes(x=participant, y=power, colour=population)) +
      theme_gar +
      geom_hline(yintercept = .83, linetype = "dashed") +
      geom_point() +
      geom_line() +
      scale_colour_viridis_d(option = "magma", 
                             begin = 0, end = 0.9) +
      scale_x_continuous(breaks = np.seq) +
      scale_y_continuous(breaks = seq(0.1,0.9,0.1)) +
      labs(x = "Number of participants", 
           y = "Power",
           colour = "Population size") +
      guides(colour = guide_legend(position = "inside")) +
       theme(legend.key.width = unit(1.5,"cm"),
            legend.position.inside = c(0.8,0.3),
            legend.title = element_text(size=16),
            legend.text = element_text(size = 14))
p.biasC <- p
p
```

#### Sample size needed to reach 83% power

```{r}
res <- apply(sim.res.pval <= 0.05, c(2,3), mean)
```

Number of participants needed to reach `r 100*targpow`% power when the population size is 50 = `r ceiling(approx(x = res[,which(npop.seq==50)], y = np.seq, xout = targpow)$y)`. 

When the population size is 100 = `r ceiling(approx(x = res[,which(npop.seq==100)], y = np.seq, xout = targpow)$y)`.  

When the population size is 300 = `r ceiling(approx(x = res[,which(npop.seq==300)], y = np.seq, xout = targpow)$y)`.

## Power for the full sample size

Same as above but we use all the participants -- no downsampling into smaller populations.

```{r, eval=FALSE}
diff <- tmres[,2]-tmres[,1]
set.seed(22222)
nsim <- 20000
np.seq <- seq(10, 120, 10) # number of participants
maxNP <- max(np.seq)

# group results
sim.res.full.pval <- matrix(NA, nrow = nsim, ncol = length(np.seq))

for(iter in 1:nsim){
  
  if(iter %% 1000 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }

    # sample participants with replacement from population
    samp.max <- sample(diff, maxNP, replace=TRUE) 
    
    for(P in 1:length(np.seq)){ # for each sample size
      sim.res.full.pval[iter, P] <- trimpval(samp.max[1:np.seq[P]], tr=0.2, null.value=60)
    } # participants
} # simulation iterations

beep(8)

save(
  nsim,
  np.seq,
  sim.res.full.pval,
  file = "./data/flp_sim_res_bias3.RData"
)
```

Get simulation results
```{r}
load("./data/flp_sim_res_bias3.RData")
res2 <- apply(sim.res.full.pval <= 0.05, 2, mean)
```

Number of participants needed to reach `r 100*targpow`% power for the full population size = `r ceiling(approx(x = res2, y = np.seq, xout = targpow)$y)`. 

## Summary figure
```{r, eval = FALSE}

plots <- align_plots(p.biasA, p.biasB, align = 'v', axis = 'l')

p.bias.bottom <- cowplot::plot_grid(plots[[2]], # p.biasB
                   p.biasC,
                   labels = c("B", "C"),
                   ncol = 2,
                   nrow = 1,
                   label_size = 20,
                   align = 'v',
                   axis = 'l',
                   hjust = -0.5,
                   scale = 1)

cowplot::plot_grid(plots[[1]], # p.biasA 
                   p.bias.bottom,
                   labels = c("A", ""),
                   rel_widths = c(1, 1), 
                   ncol = 1,
                   nrow = 2,
                   label_size = 20)

# save figure
ggsave(filename=('./figures/fig_flp_simbias.tiff'),width=13,height=10,dpi=300)
```
