---
title: "Sampling distributions from a lognormal population distribution"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

# Dependencies
```{r message=FALSE, warning=FALSE}
library(ggplot2)
source("./functions/theme_gar.txt") # define ggplot theme
library(tibble)
library(cowplot)
# library(Cairo)
```

```{r}
sessionInfo()
```

# Population
We define a lognormal population.

```{r}
x <- seq(0, 5, 0.01)
meanlog <- 0
sdlog <- 0.5
pop.dens <- dlnorm(x, meanlog = meanlog, sdlog = sdlog)
pop.m <- exp(meanlog + sdlog^2 / 2) # population mean
pop.sd <- sqrt( (exp(sdlog^2)-1)*exp(2*meanlog + sdlog^2) ) # pop sd
# exp(meanlog) # pop median
# exp(meanlog - sdlog^2) # pop mode
# test using finite population
# pop <- rlnorm(1000000, meanlog, sdlog)
# mean(pop)
# sd(pop)

# SEM
# round(pop.sd / sqrt(10), digits = 3)
# round(pop.sd / sqrt(20), digits = 3)
# round(pop.sd / sqrt(50), digits = 3)

df <- tibble(x = x, y = pop.dens)
p <- ggplot(df, aes(x=x, y=y)) + theme_gar +
 geom_line(size = 1.5) +
  geom_vline(xintercept = pop.m) +
  scale_x_continuous(breaks = seq(0, 5, .5)) +
  labs(x = "Values", y = "Density") +
  # SEM labels
  geom_label(aes(x = 3, y = 0.75, 
             label = "SEM(n=10) = 0.191"), 
             size = 7
            ) + 
  geom_label(aes(x = 3, y = 0.60, 
             label = "SEM(n=20) = 0.135"), 
             size = 7
            ) + 
  geom_label(aes(x = 3, y = 0.45, 
             label = "SEM(n=50) = 0.085"), 
             size = 7
            ) + 
  ggtitle(paste0("Lognormal distribution (\U03BC=",round(pop.m,digits = 2),", \U03C3=",round(pop.sd, digits = 3),")"))
p
pA <- p
```

For a sample of n=10 observations, the standard error of the mean (SEM) is $\frac{sd}{\sqrt{n}$} = `r round(pop.sd/sqrt(10), digits = 3)`.

# Random samples
We draw random samples of size n = 10 from the population. Because of random sampling, the sample means can be close or far from the population mean. 


```{r}
set.seed(1) # reproducible result
nsamp <- 10 # number of samples
nt <- 10 # sample size: number of trials
samp <- matrix(rlnorm(nsamp * nt, meanlog, sdlog), nrow = nsamp)

df <- tibble(val = as.vector(samp),
             Samples = rep(1:nsamp, nt))

df2 <- tibble(val = apply(samp, 1, mean),
              Samples = 1:nsamp)

# illustrate sample
p <- ggplot(df, aes(x = val, y = Samples)) + theme_gar +
  geom_jitter(height = .15, alpha = 1,
              shape = 21, fill = "grey", colour = "black") +
  geom_segment(data = df2, 
               aes(x=val, xend=val, y=Samples-0.3, yend=Samples+0.3),
               size = 0.75) +
  geom_vline(xintercept = pop.m) +
  theme(panel.grid.minor.y = element_blank()) +
  coord_cartesian(xlim = c(0, 4)) +
  scale_x_continuous(breaks = seq(0, 5, .5)) +
  scale_y_continuous(breaks = seq(1, 10, 1)) +
  labs(x = "Values from independent trials",
       y = "Random samples") +
    ggtitle("Random samples of n = 10 trials")
p
pB <- p
```

# Sampling distributions

The variability demonstrated in the previous section depends on sample size: with larger samples, the sample means tend to be closer to the population mean on average. Here we illustrate distributions of sample means, for 20000 simulations / experiments, in which we draw samples of different sizes. These distributions are called **sampling distributions**.

## Simulation
```{r}
nsim <- 20000 # number of iterations
n <- 10
sampd10 <- apply(matrix(rlnorm(n*nsim, meanlog, sdlog), nrow = nsim), 1, mean)
n <- 20
sampd20 <- apply(matrix(rlnorm(n*nsim, meanlog, sdlog), nrow = nsim), 1, mean)
n <- 50
sampd50 <- apply(matrix(rlnorm(n*nsim, meanlog, sdlog), nrow = nsim), 1, mean)
```

## Standard error of the mean (SEM)
SEM is defined as the standard deviation of the sampling distribution of the sample mean.
```{r}
sd(sampd10) # 0.1896274
sd(sampd20) # 0.1352282
sd(sampd50) # 0.08564219
```
These values, should be very similar to the values obtained using the population standard deviation --- this is the case:
```{r}
pop.sd / sqrt(10)
pop.sd / sqrt(20)
pop.sd / sqrt(50)
```
The two sets of values will converge with an increasingly large number of simulations.

## Illustrate results
```{r}
# distributions
df <- tibble(obs = c(sampd10, sampd20, sampd50),
             cond = factor(c(rep(10, nsim),
                             rep(20, nsim),
                             rep(50, nsim)))
             )

# add labels with sd of each sampling distribution
df.lab <- tibble(label = c(paste0("sd = ",round(sd(sampd10),digits = 3)), 
                           paste0("sd = ",round(sd(sampd20), digits = 3)), 
                           paste0("sd = ",round(sd(sampd50), digits = 3))),
                 cond = factor(c(10, 20, 50)),
                 x = c(1.7, 1.7, 1.7),
                 y = c(1.5, 2, 4))

# change facet labels
labels <- c("10" = "n = 10", "20" = "n = 20", "50" = "n = 50")

p <- ggplot(df, aes(x=obs)) + theme_gar +
  geom_line(stat = "density", size = 1) +
  facet_grid(cols = vars(cond), labeller=labeller(cond = labels)) +
  labs(x = "Sample means", y = "Density") + 
  geom_label(data = df.lab, 
             aes(x = x, y = y, label = label),
             size = 7) +
  ggtitle("Sampling distributions")
p
pC <- p
```

# Summary figure
```{r, eval=FALSE, warning=FALSE}
require(Cairo)
CairoPDF(file=('./figures/figure_lognormal.pdf'),width=15,height=10)
# combine panels into one figure
pAB <- cowplot::plot_grid(pA, pB,
                          labels = c("A", "B"),
                          ncol = 2,
                          nrow = 1,
                          rel_widths = c(1, 1), 
                          label_size = 20, 
                          # hjust = -0.5, 
                          scale = 1,
                          align = 'h')

cowplot::plot_grid(pAB, pC,
                          labels = c("", "C"),
                          ncol = 1,
                          nrow = 2,
                          rel_heights = c(1, 1), 
                          label_size = 20, 
                          # hjust = -0.5, 
                          scale = 1)
# save figure
# ggsave(filename=('./figures/figure_lognormal.pdf'),width=15,height=10)
dev.off()
```

