---
title: "Sampling distributions for reaction time data"
author: "Guillaume A. Rousselet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: no
    number_sections: no
    toc: yes
    toc_depth: 2
---

# Dependencies
```{r message=FALSE, warning=FALSE}
library(ggplot2)
source("./functions/theme_gar.txt") # define ggplot theme
source("./functions/trimpval.txt") # get one-sample t-test on trimmed means
source("./functions/functions.txt")
library(tibble)
# library(cowplot)
# install.packages("devtools")
# devtools::install_github("GRousselet/rogme")
library(rogme)
library(beepr)
# devtools::install_github("zeehio/facetscales") 
library(facetscales) # set different scales for each facet
```

```{r}
sessionInfo()
```

Before we conduct an experiment, we decide on a number of trials per condition, and a number of participants, and then we hope that whatever we measure comes close to the population values we’re trying to estimate. Using a large dataset we can perform data-driven simulations to illustrate how close we can get to the truth.

We use data from the [French Lexicon Project](https://sites.google.com/site/frenchlexicon/results).
Click on "French Lexicon Project trial-level results with R scripts.zip".
The `.RData` dataset was created by applying the script `getflprtdata.Rmd` available on [GitHub](https://github.com/GRousselet/rogme/tree/master/data-raw).

```{r}
# get data - tibble = `flp`
flp <- rogme::flp
# columns =
#1 = participant
#2 = rt
#3 = acc = accuracy 0/1
#4 = condition = word/non-word

# Determine population trimmed mean of trimmed mean differences
tmp <- tapply(flp$rt, list(flp$participant, flp$condition), mean, trim = 0.2)
pop.tm.diff <- mean(tmp[,2]-tmp[,1], trim = 0.2)
pop.np <- length(unique(flp$participant))
```

N = `r length(unique(flp$participant))` participants.

# Illustrate random sample of participants

## Create data frame
```{r}
set.seed(3)
# participants' indices
p.list <- as.numeric(as.character(unique(flp$participant)))
np <- 20
rsamp <- sample(p.list, np, replace = FALSE)
df <- flp[flp$participant %in% rsamp, ]
df$participant <- as.character(df$participant)
df$participant <- factor(df$participant, levels=unique(df$participant))
#levels(df$condition)
levels(df$condition) <- c("Word", "Non-Word")
```

## Make figure
```{r, fig.height=4, fig.width=10}
p <- ggplot(df, aes(x = rt)) + theme_gar +
  geom_line(stat = "density", aes(colour=participant), size = 0.5) + 
  scale_color_viridis_d() + 
  scale_x_continuous(breaks=seq(0,2000,500), minor_breaks = seq(0, 2000, 100)) +
  coord_cartesian(xlim = c(0, 2000)) +
  theme(legend.position = "none") + 
  labs(x = "Reaction times (ms)", y = "Density") +
  facet_grid(. ~ condition)
p
p.20 <- p
```

The distributions are positively skewed, as expected for RT data, and participants tend to be slower in the non-word condition compared to the word condition. Usually, a single number is used to summarise each individual RT distribution. From 1000 values to 1, that’s some serious data compression! In psychology, the mean is often used, but here we use the 20% trimmed mean, which gives a better indication of the location of the typical observation and protects against the influence of outliers. 

# Summary results for all participants

Here is the distribution across participants of trimmed mean RT for the word and non-word conditions and their differences.

## Compute trimmed means
```{r}
# get data: mean RT for every participant
tmres <- tapply(flp$rt, list(flp$participant, flp$condition), mean, trim = 0.2)
summary(tmres)
```

## Illustrate results

All distributions are skewed.

```{r, fig.height=4, fig.width=10}

# create data frame
df <- tibble(x = c(as.vector(tmres),tmres[,2]-tmres[,1]),
             Condition = rep(c("Word", "Non-Word", "Difference"), each = pop.np)
             )
df$Condition <- as.character(df$Condition)
df$Condition <- factor(df$Condition, levels=unique(df$Condition))

scales_x <- list(
  `Word` = scale_x_continuous(limits=c(0,1600), breaks=seq(0,1600,250)),
  `Non-Word` = scale_x_continuous(limits=c(0,1600), breaks=seq(0,1600,250)),
  `Difference` = scale_x_continuous(limits=c(-100,500), breaks=seq(-100,500,100))
)

# make plot
p <- ggplot(df, aes(x = x)) + theme_gar + 
  geom_line(stat = "density", size = 1.5) + 
  labs(x = "20% trimmed mean reaction times (ms)", y = "Density") +
  facet_grid_sc(. ~ Condition, scales = list(x = scales_x))
p
p.alltm <- p
```

`r round(mean(tmres[,2]-tmres[,1]>0)*100, digits=1)`% of participants had a positive difference, meaning that they tended to be faster in the Word than the Non-Word condition.

# Hierarchical sampling: effect

We sample participants and trials with replacement. For each condition, participant and trial sample size, we compute the 20% trimmed mean. Then, for each condition and participant sample size we compute the group 20% trimmed mean and the *p* value based on a t-test for trimmed means.

```{r, eval = FALSE}
set.seed(22222)
p.list <- as.numeric(as.character(unique(flp$participant)))
nsim <- 5000
np.seq <- c(10, 20, 50) # number of participants
maxNP <- max(np.seq)
nt.seq <- seq(50, 250, 50) # number of trials
maxNT <- max(nt.seq)

# intermediate results with single trials
diff.all <- matrix(NA, nrow = maxNP, ncol = length(nt.seq))

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(nt.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  if(iter %% 100 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  # Sample participants with replacement
  bootid <- sample(p.list, maxNP, replace = TRUE)
  # Sample trials with replacement 
  for(P in 1:maxNP){ # for each bootstrap participant
    boot.w <- sample(flp$rt[flp$participant %in% bootid[P] & flp$condition == "word"],
      maxNT, replace = TRUE)
    boot.nw <- sample(flp$rt[flp$participant %in% bootid[P] & flp$condition == "non-word"],
      maxNT, replace = TRUE)
    
    # compute individual differences between trimmed means
    for(T in 1:length(nt.seq)){ # for each sample size
      # compute trimmed means
      diff.all[P,T] <- mean(boot.nw[1:nt.seq[T]], trim = 0.2) - mean(boot.w[1:nt.seq[T]], trim = 0.2)
    }
  }
  
  # compute group trimmed means and p values 
  for(NP in 1:length(np.seq)){
    for(NT in 1:length(nt.seq)){
    sim.res[iter, NP, NT] <- mean(diff.all[1:np.seq[NP],NT], trim = 0.2)
    sim.res.pval[iter, NP, NT] <- trimpval(diff.all[1:np.seq[NP],NT], tr = 0.2)
    }
  }
}

save(
  nsim,
  np.seq,
  nt.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res.RData"  
)

beep(8)
```

## Illustrate results

### Panel 1
```{r, fig.height=4, fig.width=10}
load(file = "./data/flp_sim_res.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
             )

levels(df$np) <- c("10 participants", "20 participants", "50 participants")

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = pop.tm.diff) +
  geom_line(stat = "density", size = 0.75) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p.diff1 <- p
p
```

### Panel 2
```{r, fig.height=4, fig.width=10}
alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res[v.sim.res.pval>alpha] <- NA

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
             )

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = pop.tm.diff) +
  geom_line(stat = "density", size = 0.75) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

The vertical line in each panel marks the population 20% trimmed mean.
Here, conditioning on $p\leq0.05$ has very little effect on the sampling distributions because the population effect size is so large.

# Hierarchical sampling: no effect

In this situation, trials are sampled with replacement from the two conditions merged together, so that on average, there is no effect.

```{r, eval = FALSE}
set.seed(22222)
p.list <- as.numeric(as.character(unique(flp$participant)))
nsim <- 5000
np.seq <- c(10, 20, 50) # number of participants
maxNP <- max(np.seq)
nt.seq <- seq(50, 250, 50) # number of trials
maxNT <- max(nt.seq)

# intermediate results with single trials
diff.all <- matrix(NA, nrow = maxNP, ncol = length(nt.seq))

# group results
sim.res <- array(NA, dim = c(nsim, length(np.seq), length(nt.seq))) 
sim.res.pval <- sim.res

for(iter in 1:nsim){
  if(iter %% 100 == 0){
    print(paste("Simulation ",iter," / ",nsim,"..."))
    beep(2)
  }
  # Sample participants with replacement
  bootid <- sample(p.list, maxNP, replace = TRUE)
  # Sample trials with replacement 
  for(P in 1:maxNP){ # for each bootstrap participant
    all.trials <- flp$rt[flp$participant %in% bootid[P] & 
                        (flp$condition == "word" | flp$condition == "non-word")]
    boot.w <- sample(all.trials, maxNT, replace = TRUE)
    boot.nw <- sample(all.trials, maxNT, replace = TRUE)
    
    # compute individual differences between trimmed means
    for(T in 1:length(nt.seq)){ # for each sample size
      # compute trimmed means
      diff.all[P,T] <- mean(boot.nw[1:nt.seq[T]], trim = 0.2) - mean(boot.w[1:nt.seq[T]], trim = 0.2)
    }
  }
  
  # compute group trimmed means and p values 
  for(NP in 1:length(np.seq)){
    for(NT in 1:length(nt.seq)){
    sim.res[iter, NP, NT] <- mean(diff.all[1:np.seq[NP],NT], trim = 0.2)
    sim.res.pval[iter, NP, NT] <- trimpval(diff.all[1:np.seq[NP],NT], tr = 0.2)
    }
  }
}

save(
  nsim,
  np.seq,
  nt.seq,
  sim.res,
  sim.res.pval,
  file = "./data/flp_sim_res_noeffect.RData"  
)

beep(8)
```

## Illustrate results

### Panel 1
```{r, fig.height=4, fig.width=10}
load(file = "./data/flp_sim_res_noeffect.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
             )

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = 0) + # long-run effect size
  geom_line(stat = "density", size = 0.75) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

### Panel 2
```{r, fig.height=4, fig.width=10}
alpha <- 0.05 # arbitrary cut-off

v.sim.res <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res[v.sim.res.pval>alpha] <- NA

df <- tibble(x = v.sim.res,
             np = factor(rep(rep(np.seq, each = nsim), length(nt.seq))),
             nt = factor(rep(nt.seq, each = nsim * length(np.seq)))
             )

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = pop.tm.diff) +
  geom_line(stat = "density", size = 0.75) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p
```

Results conditioned on $p\leq0.05$ are symmetric around zero. With 10 participants and 50 trials per condition, the median of the absolute group differences is `r round(median(abs(df$x[df$np=="10" & df$nt=="50"]), na.rm = TRUE), digits=1)` ms, and among the experiments reporting significant results, 10% report differences at least as large as `r round(quantile(abs(df$x[df$np=="10" & df$nt=="50"]), na.rm = TRUE, probs=0.9, type=8), digits=1)` ms. With 50 participants and 200 trials per condition, the median of the absolute group differences is `r round(median(abs(df$x[df$np=="50" & df$nt=="200"]), na.rm = TRUE), digits=1)` ms, and among the experiments reporting significant results, 10%  report differences at least as large as `r round(quantile(abs(df$x[df$np=="50" & df$nt=="200"]), na.rm = TRUE, probs=0.9, type=8), digits=1)` ms.

### Merged panels using facetting
```{r, eval = FALSE}
load(file = "./data/flp_sim_res_noeffect.RData")

alpha <- 0.05 # arbitrary cut-off

v.sim.res.raw <- as.vector(sim.res)
v.sim.res.cond <- as.vector(sim.res)
v.sim.res.pval <- as.vector(sim.res.pval)
v.sim.res.cond[v.sim.res.pval>alpha] <- NA

df <- tibble(x = c(v.sim.res.raw, v.sim.res.cond),
             np = factor(rep(rep(rep(np.seq, each = nsim), length(nt.seq)),2)),
             nt = factor(rep(rep(nt.seq, each = nsim * length(np.seq)),2)),
             cond = factor(rep(c("Raw data","Cond. data"), 
                                each = nsim * length(np.seq) * length(nt.seq)))
             )

levels(df$np) <- c("10 participants", "20 participants", "50 participants")
df$cond <- keeporder(df$cond)

p <- ggplot(df, aes(x = x, colour = nt)) + theme_gar +
  geom_vline(xintercept = 0) + # long-run effect size
  geom_line(stat = "density", size = 0.75) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Number of trials")) +
  scale_colour_viridis_d(option = "plasma", 
                         begin = 0, end = 0.7) + 
  facet_grid(rows = vars(cond), cols = vars(np), scales = "fixed") +
  labs(x = "Group 20% trimmed mean difference", y = "Density")
p.diff.noeffect <- p
p
```

# Summary figure
```{r, eval = FALSE}

cowplot::plot_grid(p.20, 
                   p.alltm, 
                   p.diff1, 
                   p.diff.noeffect + theme(legend.position = "none"),
                    labels = c("A", "B", "C", "D"),
                    ncol = 1,
                    nrow = 4,
                    rel_heights = c(1, 0.75, 0.75, 1), 
                    label_size = 20,
                    align = 'v',
                    axis = 'l',
                    hjust = -0.5,
                    scale = 1)

# save figure
ggsave(filename=('./figures/figure_flp.pdf'),width=13,height=20)
```

